---
title: "Okuyama Mechanistic Model Report"
subtitle: "Mechanistic Functional Response Analysis"
author: "Segundo Núñez-Campero"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1]{fontenc}
  - \usepackage{float}
params:
  DATASET: NULL
  PARAMS_ALL: NULL
  PROFILES_ALL: NULL
  CURVE_ALL: NULL
  HIST_ALL: NULL
  AIC_TABLE: NULL
  Z_CI: NULL
  REPORT_ELEMENTS: NULL
  INCLUDE_DATASET: FALSE
  T_EXP: NULL
  ITERMAX: NULL
  NP: NULL
  RELTOL: NULL
  N_SIM: NULL
  Z_MIN: NULL
  Z_MAX: NULL
  Z_STEP: NULL
  CORES: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

# Fallback para params
if (!exists("params")) {
  params <- list(
    DATASET = NULL,
    PARAMS_ALL = NULL,
    PROFILES_ALL = NULL,
    CURVE_ALL = NULL,
    HIST_ALL = NULL,
    AIC_TABLE = NULL,
    Z_CI = NULL,
    REPORT_ELEMENTS = NULL,
    INCLUDE_DATASET = FALSE
  )
}
```

# Cover Page

```{r cover, results='asis', echo=FALSE}
if (knitr::is_latex_output()) {
  cat('
\\begin{center}
{\\huge \\bfseries Okuyama Mechanistic Model Report} \\\\[0.5cm]
{\\large Mechanistic Functional Response Analysis} \\\\[1cm]
{\\large \\textbf{Author:} Segundo Núñez-Campero} \\\\
{\\large \\textbf{Date:}', format(Sys.time(), '%B %d, %Y'), '} \\\\
{\\large \\textbf{Model:} Okuyama (2012), Biological Control DOI <10.1016/j.biocontrol.2011.10.008>}
\\end{center}
\\vspace{1cm}
\\hrule
\\vspace{0.5cm}
')
} else {
  cat('<div style="padding:20px; border:2px solid #333; border-radius:10px; background:#f7f7f7;">
<h2>Okuyama Mechanistic Model Report</h2>
<h4>Mechanistic Functional Response Analysis</h4>
<p><strong>Author:</strong> Segundo Núñez-Campero</p>
<p><strong>Date:</strong>', format(Sys.time(), '%B %d, %Y'), '</p>
<p><strong>Model:</strong> Okuyama (2012), Biological Control DOI <10.1016/j.biocontrol.2011.10.008</p>
</div>')
}
```

# Dataset (optional)

```{r dataset}
if (isTRUE(params$INCLUDE_DATASET) && !is.null(params$DATASET)) {
  kable(head(params$DATASET), caption = "First rows of the dataset used in the analysis") %>%
    kable_styling(latex_options = "hold_position")
}
```

# Summary Statistics

```{r summary}
if (!is.null(params$DATASET)) {
  df <- params$DATASET

  summary_table <- df %>%
    group_by(species) %>%
    summarise(
      n = n(),
      mean_dens = mean(dens),
      mean_par = mean(par),
      sd_par = sd(par),
      min_par = min(par),
      max_par = max(par),
      .groups = "drop"
    )

  kable(summary_table, caption = "Summary statistics by species") %>%
    kable_styling(latex_options = "hold_position")
}
```

# Fitted Parameters

```{r params}
if (!is.null(params$PARAMS_ALL) && "params" %in% params$REPORT_ELEMENTS) {
  kable(params$PARAMS_ALL, caption = "Fitted parameters for each species") %>%
    kable_styling(latex_options = "hold_position")
}
```

# Interpretation of Parameters

```{r interpretation}
if (!is.null(params$PARAMS_ALL)) {

  interpret <- function(a, h, z, k, s) {
    out <- c()

    if (z < 1) out <- c(out, "Sublinear response: weak acceleration with density.")
    if (z >= 1 && z < 1.2) out <- c(out, "Approximately Type II response.")
    if (z >= 1.2) out <- c(out, "Type III-like response: strong acceleration at low density.")

    if (h < 0.05) out <- c(out, "Very fast handling time.")
    if (h >= 0.05 && h < 0.15) out <- c(out, "Moderate handling time.")
    if (h >= 0.15) out <- c(out, "Slow handling time: strong saturation.")

    if (s > 0.2) out <- c(out, "High variability in handling time.")
    if (k < 1) out <- c(out, "High variability in search time.")

    paste(out, collapse = " ")
  }

  interp_table <- params$PARAMS_ALL %>%
    rowwise() %>%
    mutate(interpretation = interpret(a, h, z, k, s)) %>%
    ungroup()

  kable(interp_table, caption = "Automatic interpretation of fitted parameters") %>%
    kable_styling(latex_options = "hold_position")
}
```

# Confidence Intervals for z (Profile Likelihood)

```{r z_ci}
if (!is.null(params$Z_CI)) {
  kable(params$Z_CI, caption = "95% Confidence intervals for z (profile likelihood)") %>%
    kable_styling(latex_options = "hold_position")
}
```

# Model Comparison (AIC)

```{r aic}
if (!is.null(params$AIC_TABLE)) {
  # Añadir columna de interpretación de delta AIC
  aic_table <- params$AIC_TABLE %>%
    mutate(
      support = case_when(
        is.na(delta_AIC) ~ "Not computed",
        delta_AIC > 10   ~ "Strong support for full model (z free)",
        delta_AIC > 2    ~ "Moderate support for full model",
        delta_AIC >= -2  ~ "Equivalent support",
        delta_AIC < -2   ~ "Support for restricted model (z=1)"
      )
    )
  kable(aic_table, caption = "AIC comparison between full model (z free) and restricted model (z=1)") %>%
    kable_styling(latex_options = "hold_position")
}
```

# Likelihood Profile for z

```{r profile}
if (!is.null(params$PROFILES_ALL) && "profile" %in% params$REPORT_ELEMENTS) {
  p_prof <- ggplot(params$PROFILES_ALL, aes(x = z, y = nll, color = species)) +
    geom_line(size = 1) +
    theme_bw() +
    labs(title = "Likelihood profile for z",
         x = "z",
         y = "Negative log-likelihood")
  print(p_prof)
}
```

# Stochastic Mechanistic Curve

```{r curve}
if (!is.null(params$CURVE_ALL) && "curve" %in% params$REPORT_ELEMENTS) {
  p_curve <- ggplot(params$CURVE_ALL, aes(x = dens, y = mean, color = species)) +
    geom_line(size = 1) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = species),
                alpha = 0.25, color = NA) +
    theme_bw() +
    labs(title = "Stochastic mechanistic curve",
         x = "Host density",
         y = "Parasitized hosts")
  print(p_curve)
}
```

# Distribution Plots

## Histograms

```{r hist}
if (!is.null(params$HIST_ALL) && "hist" %in% params$REPORT_ELEMENTS) {
  p_hist <- ggplot(params$HIST_ALL, aes(x = paras)) +
    geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
    facet_grid(species ~ dens) +
    theme_bw() +
    labs(title = "Histogram of simulated parasitism",
         x = "Parasitized hosts", y = "Count")
  print(p_hist)
}
```

## Kernel Density

```{r kernel}
if (!is.null(params$HIST_ALL) && "kernel" %in% params$REPORT_ELEMENTS) {
  p_kernel <- ggplot(params$HIST_ALL, aes(x = paras, color = species, fill = species)) +
    geom_density(alpha = 0.4) +
    facet_grid(species ~ dens) +
    theme_bw() +
    labs(title = "Kernel density of simulated parasitism",
         x= "Parasitized hosts", y= "Density")
  print(p_kernel)
}
```

## Boxplots

```{r box}
if (!is.null(params$HIST_ALL) && "box" %in% params$REPORT_ELEMENTS) {
  p_box <- ggplot(params$HIST_ALL, aes(x = factor(dens), y = paras, fill = species)) +
    geom_boxplot() +
    theme_bw() +
    labs(title = "Boxplots of simulated parasitism",
         x = "Density x Species", y = "Parasitized hosts")
  print(p_box)
}
```

## Violin Plots

```{r violin}
if (!is.null(params$HIST_ALL) && "violin" %in% params$REPORT_ELEMENTS) {
  p_violin <- ggplot(params$HIST_ALL, aes(x = factor(dens), y = paras, fill = species)) +
    geom_violin(alpha = 0.4) +
    geom_jitter(aes(color = species),
            position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.9),
            alpha = 0.4, size = 1.5) +
    theme_bw() +
    labs(title = "Violin plots of simulated parasitism",
         x = "Density x Species", y = "Parasitized hosts")
  print(p_violin)
}
```

## Fan Plot

```{r fan}
if (!is.null(params$HIST_ALL) && "fan" %in% params$REPORT_ELEMENTS) {
  p_fan <- ggplot(params$HIST_ALL, aes(x = dens, y = paras, color = species)) +
    geom_point(alpha = 0.1) +
    geom_smooth(method = "loess", se = FALSE) +
    theme_bw() +
    labs(title = "Fan plot of simulated parasitism",
         x = "Host density", y = "Parasitized hosts")
  print(p_fan)
}
```

# Comparison Between Species

```{r comparison}
if (!is.null(params$PARAMS_ALL) && length(unique(params$PARAMS_ALL$species)) >= 2) {

  comp <- params$PARAMS_ALL %>%
    select(species, a, h, z, k, s) %>%
    pivot_longer(cols = -species, names_to = "parameter", values_to = "value")

  p_comp <- ggplot(comp, aes(x = species, y = value, fill = species)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~parameter, scales = "free") +
    theme_bw() +
    labs(title = "Comparison of parameter values between species",
         y = "Parameter value")

  print(p_comp)
}
```

# Analysis settings

```{r analysis-settings, echo=FALSE}
settings <- data.frame(
  Parameter = c("Exposure time (T_exp)",
                "Max iterations (itermax)",
                "Population size (NP)",
                "Relative tolerance (reltol)",
                "Number of simulations (n_sim)",
                "Min z",
                "Max z",
                "Step z",
                "CPU cores"),
  Value = c(
    params$T_EXP,
    params$ITERMAX,
    params$NP,
    params$RELTOL,
    params$N_SIM,
    params$Z_MIN,
    params$Z_MAX,
    params$Z_STEP,
    params$CORES
  )
)

knitr::kable(settings, caption = "Analysis settings used in the Shiny app.") %>%
    kableExtra::kable_styling(latex_options = "hold_position")
```


# Notes

This report was generated automatically using the Shiny application developed by Segundo Núñez-Campero, implementing the mechanistic model of Okuyama (2012).
